Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario

### 2021-04-28 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.Rmd
V1 Version for submission to Social Science & Medicine

---
title: "Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario"
author:
  - name: Author One
    email: author1@example.com
    affiliation: Some University
  - name: Author Two
    email: author2@example.com
    affiliation: Another University
    footnote: 1
  - name: Author Three
    email: author3@example.com
    affiliation: Some University
  - name: Author Four
    email: author4@example.com
    affiliation: Some University
address:
  - code: Some University
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  In this paper we analyze the changes in accessibility to food banks and related services before and during the COVID-19 pandemic in the City of Hamilton, Ontario. Food banks and services are the last line of support for households facing food insecurity; as such, their relevance cannot be ignored in the midst of the economic upheaval caused by the pandemic. Our analysis is based on the application of balanced floating catchment areas, and concentrates on households with lower incomes (<CAD40,000, approximately the Low Income Cutoff Value for a city of Hamilton's size). We find that accessibility was low to begin with in suburban and exurban parts of the city; furthermore, about 14\% of locations originally available in Hamilton closed during the pandemic, further reducing accessibility. The impact of closures on the level of service of the remaining facilities, and on accessibility, was disproportionate, with system-wide losses exceeding 40\%. Those losses were geographically and demographically uneven. While every part of the city faced a reduction in accessibility, inner suburbs fared worse in terms of loss of accessibility. As well, children (age $\le 18$) appear to have been impacted the most.

journal: "Social Science and Medicine"
date: "`r Sys.Date()`"
bibliography: references.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
header-includes:
    - \usepackage[table]{xcolor}
---

# CRediT author statement

**Author 1:** Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data Curation, Writing - Original Draft, Visualization, Supervision; **Author 2:** Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data Curation, Writing - Original Draft, Visualization; **Author 3:** Conceptualization, Investigation, Data Curation, Writing - Original Draft; **Author 4:** Resources, Writing - Original Draft, Supervision

@@

\newpage

@@

In the absence of information regarding how food insecure households travel to food banks and related services, we examined accessibility to food banks using a 15 minutes (one way) travel time threshold. The fact that we must rely on a standard created for accessible drinking water in the developing world only serves to highlight the tragedy of food insecurity in an affluent country like Canada. More broadly, it points to the absurd need to understand how a bad situation was made worse by the pandemic: in effect, the analysis reveals that disparities in accessibility to food banks and services predated the pandemic, that the pandemic contributed to the deterioration of these services, and that populations already in distress, particularly children, ended up in an even more adverse state. How much worse, it is impossible to say, mainly because there is also a dearth of information, let alone standards, regarding what is an _acceptable_ or _sufficient_ level of service when it comes to food banks. Whether this is one facility per 1,000 population or three facilities per 1,000 population, or any other number, we know not. But like international human rights standards for accessibility to clean drinking water within 30 minutes (round-trip), this work strongly suggests that adopting better defined standards for access to food banks that move beyond generalized conceptualizations of "access" to consider the geographic component of accessibility (and its interactions with travel behaviour and individual/household socioeconomic and demographic traits) would make a meaningful contribution to understanding food access and "food bank deserts" for food insecure populations.

@@

In terms of future research, it is important to note that the degree to which our low-income cut-off of CAD\$40,000 reflects food insecurity in the different zones of the study area is not known. We also consider all food banks and food service locations equally. Information on differential capacities at food provider locations is currently not collected, and given the closure of facilities was not possible to obtain. Data on services, such as the number of meals served, could be used to refine the analysis in the future. Moreover, while the travel survey allows us to model multi-modal accessibilities that align with travel behaviour observed in the travel survey and capture differences in accessibility by age categories, the use of travel survey data for modelling food insecurity also has its limitations. Research into the population weighting methods used in the TTS note that the survey may under-count the lowest- and highest-income households in the survey study region, although the magnitude of this under-counting is unknown, and approximately 20% of respondents to the 2016 survey did not report their income [@rose2018transportation]. In that regard, the modal splits of low-income households observed through the TTS data may not accurately reflect the travel behaviours of food insecure populations, and our estimates might be somewhat conservative if those households who do not report income rely more on walking and/or transit for their mobility needs.

>>>>

---
title: "Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario"
author:
  - name: Author One
    email: author1@example.com
    affiliation: Some University
  - name: Author Two
    email: author2@example.com
    affiliation: Another University
    footnote: 1
  - name: Author Three
    email: author3@example.com
    affiliation: Some University
  - name: Author Four
    email: author4@example.com
    affiliation: Some University
address:
  - code: Some University
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  In this paper we analyze the changes in accessibility to food banks and related services before and during the COVID-19 pandemic in the City of Hamilton, Ontario. Food banks and services are the last line of support for households facing food insecurity; as such, their relevance cannot be ignored in the midst of the economic upheaval caused by the pandemic. Our analysis is based on the application of balanced floating catchment areas, and concentrates on households with lower incomes (<CAD40,000, approximately the Low Income Cutoff Value for a city of Hamilton's size). We find that accessibility was low to begin with in suburban and exurban parts of the city; furthermore, about 14\% of locations originally available in Hamilton closed during the pandemic, further reducing accessibility. The impact of closures on the level of service of the remaining facilities, and on accessibility, was disproportionate, with system-wide losses exceeding 40\%. Those losses were geographically and demographically uneven. While every part of the city faced a reduction in accessibility, inner suburbs fared worse in terms of loss of accessibility. As well, children (age $\le 18$) appear to have been impacted the most.
journal: "Social Science and Medicine"
date: "`r Sys.Date()`"
bibliography: references.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
header-includes:
    - \usepackage[table]{xcolor}
---

@@

@@

In the absence of information regarding how food insecure households travel to food banks and related services, we examined accessibility to food banks using a 15 minutes (one-way) travel time threshold. The fact that we must rely on a standard created for accessible drinking water in the developing world only serves to highlight the tragedy of food insecurity in an affluent country like Canada. More broadly, it points to the absurd need to understand how a bad situation was made worse by the pandemic: in effect, the analysis reveals that disparities in accessibility to food banks and services predated the pandemic, that the pandemic contributed to the deterioration of these services, and that populations already in distress, particularly children, ended up in an even more adverse state. How much worse, it is impossible to say, mainly because there is also a dearth of information, let alone standards, regarding what is an _acceptable_ or _sufficient_ level of service when it comes to food banks. Whether this is one facility per 1,000 population or three facilities per 1,000 population, or any other number, we know not. But like international human rights standards for accessibility to clean drinking water within 30 minutes (round-trip), this work strongly suggests that adopting better defined standards for access to food banks that move beyond generalized conceptualizations of "access" to consider the geographic component of accessibility (and its interactions with travel behaviour and individual/household socioeconomic and demographic traits) would make a meaningful contribution to understanding food access and "food bank deserts" for food insecure populations.

@@

In terms of future research, it is important to note that the degree to which our low-income cut-off of CAD\$40,000 reflects food insecurity in the different zones of the study area is not known. We also consider all food banks and food service locations equally. Information on differential capacities at food provider locations is currently not collected, and given the closure of facilities was not possible to obtain. Data on services, such as the number of meals served, could be used to refine the analysis in the future. Moreover, while the travel survey allows us to model multi-modal accessibilities that align with travel behaviour observed in the travel survey and capture differences in accessibility by age categories, the use of travel survey data for modelling food insecurity also has its limitations. Research into the population weighting methods used in the TTS note that the survey may under-count the lowest- and highest-income households in the survey study region, although the magnitude of this under-counting is unknown, and approximately 20% of respondents to the 2016 survey did not report their income [@rose2018transportation]. In that regard, the modal splits of low-income households observed through the TTS data may not accurately reflect the travel behaviours of food insecure populations, and our estimates might in fact be somewhat conservative if those households who do not report income rely more on walking and/or transit for their mobility needs.

### 2021-04-28 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.tex
V1 Version for submission to Social Science & Medicine

\hypertarget{credit-author-statement}{%
\section{CRediT author statement}\label{credit-author-statement}}

\textbf{Author 1:} Conceptualization, Methodology, Software, Validation,
Formal analysis, Investigation, Data Curation, Writing - Original Draft,
Visualization, Supervision; \textbf{Author 2:} Conceptualization,
Methodology, Software, Validation, Formal analysis, Investigation, Data
Curation, Writing - Original Draft, Visualization; \textbf{Author 3:}
Conceptualization, Investigation, Data Curation, Writing - Original
Draft; \textbf{Author 4:} Resources, Writing - Original Draft,
Supervision

\newpage

@@

Food insecurity is defined as an ``inadequate or uncertain access to a
sufficient quantity and/or adequate quality of food'' due to a
household's financial limitations (Enns et al., 2020). This condition
has been associated with reductions in nutritional outcomes
(Bhattacharya et al., 2004; Kirkpatrick and Tarasuk, 2008; Olson, 1999)
and negative physical and mental health impact in children and adults
(Elgar et al., 2021; Jones, 2017; Ramsey et al., 2011; Seligman et al.,
2010; Stuff et al., 2004). Over at least the past four decades food
banks and related services have become an essential line of defense
against food insecurity in Canadian communities (Black and Seto, 2020;
Holmes et al., 2018; Riches, 2002; Tarasuk et al., 2020). In this
respect, Canada is not unlike numerous other wealthy countries where a
systematic dismantling of the welfare state took place in the
intervening period (Tarasuk et al., 2014).

@@

It is known that urban food environments, within which people make their
daily food choices, are essential in influencing eating behaviours and
health outcomes, based on factors such as food availability, ease of
geographic accessibility and socio-demographic variations (Paez et al.,
2010; Vanderlee and L'Abbé, 2017; Widener, 2018). However, while there
is a wealth of literature that has examined the topic of geographic
accessibility to healthy food through the ``food desert'' concept, there
has been little research into accessibility to food banks. Although
previous work has explored differences in \emph{accessing} food banks,
such as how some households utilize food banks over short periods of
time while others regularly utilize food banks as longer-term resource
(e.g., Enns et al., 2020), we are not aware of any research that has
focused on the geographic component of accessibility.

@@

For this study, we aim to look at how the landscape of food banks and
related services (e.g.~low-cost or free meal service providers)
available in Hamilton, Ontario, changed during the pandemic. Did the
number of open food bank services diminish? If so, what was the
accessibility to food banks before the pandemic from the perspective of
low income households, and how it changed during the pandemic? And
finally, who are most likely to have been impacted by changes in the
accessibility landscape? This paper first looks at the distribution of
food banks and related services before and during the pandemic. Then, we
use the balanced floating catchment area approach of Paez et al. (2019)
to investigate the accessibility situation. For this, we use a fully
disaggregated approach based on parcel-level data. Socio-economic and
demographic data are drawn from the latest Census of Canada (2016),
whereas travel information is from the most recent regional travel
survey from 2016. This paper follows reproducible research
recommendations (see Brunsdon and Comber, 2020), and the research was
conducted using open source tools for transportation analysis (Lovelace,
2021). The code and data necessary to reproduce the analysis are
available in a public repository\footnote{https://drive.google.com/drive/folders/1-l8hO1pMIqaBqf57j\_M\_BMrXhXFvrV2c?usp=sharing}.

@@

Food insecurity is the inability to acquire and consume an adequate
amount or good quality food, leading to inadequate nutrient intake (Enns
et al., 2020; Kirkpatrick and Tarasuk, 2008; Tarasuk and Vogt, 2009).
This nutrient deficiency is associated with major population health
concerns, particularly among Canadians at socio-economic disadvantage
(Bazerghi et al., 2016). Data on food insecurity in Canada is collected
in several ways. Quantitatively, official government surveys such as the
Household Food Security Survey Module (HFSSM), the Canadian Community
Health Surveys (CCHS), the International Study of Adults (LISA), and
official classifications determined by Health Canada in relation to
socio-demographic variables offer some insight into food insecurity
(Gundersen et al., 2018; Kirkpatrick and Tarasuk, 2008; Tarasuk and
Vogt, 2009). Nationally, it has been found that food insecurity impacts
approximately 12.3\% of Canadian households (Tarasuk et al., 2014)

@@

Economic stress caused by the COVID-19 pandemic, rising unemployment
rates, and changes in poverty levels have disrupted the food environment
and led to higher rates of food insecurity (Niles et al., 2020).
Pandemic-related food security studies in the US have found a
substantial increase in households experiencing food insecurity for the
first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days (Statistics Canada,
2020a). Considering the negative mental and physical health effects
associated with food insecurity, increases in food insecurity rates due
to the pandemic, signal a change in the food environment with potential
damages to population health during the course of the pandemic and
beyond (Niles et al., 2020). Recent data\footnote{https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report\_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis\_Exec-Sum\_EN.pdf}
from FBC showed that 52\% of member food banks reported an increase in
usage in March of 2020 when initial lockdown restrictions were put in
place across much of the country. The pandemic also created significant
staffing issues with 42\% of food banks reporting a reduction in
volunteers. However, 53\% of food banks later reported a decrease in use
into the summer of 2020 which FBC members attributed to emergency
financial support programs from the federal government. Nevertheless,
some of these benefit programs were temporary. Although more recent
statistics on food bank use in Hamilton in 2020 and 2021 are not yet
available, data from the Daily Bread Food Bank in neighbouring Toronto
for August 2020 shows visits climbing 51\% year-over-year in that
city\footnote{https://www.dailybread.ca/wp-content/uploads/2020/11/Whos-Hungry-Report-2020.pdf},
which suggests that many households in Hamilton are likely to turn to
food bank services to meet their needs.

@@

Geographically, the ``old'' city of Hamilton was amalgamated with
several of its surrounding municipalities in 2001, with the city now
featuring a mix of urban, suburban, exurban, and rural areas. Lower-cost
housing proximate to the city's industrial north end has traditionally
attracted immigrants and less-affluent residents compared to the city's
wealthier suburbs. However, the decentralization of population from the
inner core has led to challenges in transit connectivity to amenities
and services and the proportion of auto users compared to transit users
remains very high (Behan et al., 2008; Topalovic et al., 2012). In
addition, the city is separated geographically by the Niagara
Escarpment. With sections of rocky cliff that approach 100m in height,
the escarpment presents a significant challenge for promoting active
travel and transport connections between ``mountain'' and ``lower city''
neighbourhoods. Taken together, the high level of food need, population
locations, and transportation network characteristics combine to inform
spatial accessibility to food banks and food services in the city.

@@

Figure \ref{fig:sensitivity-analysis} shows the results of conducting
sensitivity analysis of the system-wide accessibility as we vary the
threshold (considering the situation before the pandemic). There is a
clear pattern whereby more strict values of \(\delta\) are associated
with higher levels of system-wide accessibility: while increases in
accessibility that result from decreases in the travel time window might
seem counter-intuitive, this is a result of lower \emph{congestion},
since fewer households are serviced and thus competition for the same
resources is more limited. System-wide accessibility declines with
higher values of \(\delta\): as more households are serviced, congestion
grows and the level of service declines, although this happens at a
declining rate. We are not aware of any research that explains how long
people are expected to travel for food banks, but we note that in
developing countries, accessible sources of drinking water are those
that can be reached in less than 30 minutes (round trip, see UNICEF-WHO,
2019). There is no reason why people in affluent countries should be
expected to spend more time travelling for a basic necessity such as
food. Accordingly, we adopt a 15-minute threshold for the analysis
(representing a one-way trip). This threshold is also approximately
where the rate of change in accessibility slows down.

@@

We find that the accessibility landscape deteriorated substantially
during the pandemic, with accessibility dropping on average by almost
41\%, but with large variations: some zones experienced changes in
accessibility of only about 10\%, whereas the most affected zone saw a
loss of accessibility of almost 71\%. Figure
\ref{fig:accessibility-changes-with-local-i} shows the changes in
accessibility. Every zone is worse off after the closure of facilities
during the pandemic, but some parts of the city seem to have been
particularly affected. To better highlight these changes, we used a
local indicator of spatial autocorrelation (Anselin, 1995) to explore
the pattern of change in accessibility. Twenty-four TAZ are flagged as
having significantly large losses of accessibility (at \(p\le 0.10\),
without correcting for multiple comparisons). Those zones are
highlighted in the figure, where it can be seen that they form more or
less compact neighborhoods. Remarkably, the largest significant drops in
accessibility are not downtown, but located in two cases in the
industrial north of the city, in one case in an inner suburb above the
escarpment, and lastly in a more suburban/exurban region in the
south-west.

@@

For the more suburban clusters of zones, the decrease in accessibility
is derived from the closure of locations throughout the city reachable
by car. In the cluster of central suburban zones for example, low-income
households in the outer ring of zones that exhibit medium to high
decreases in accessibility within this cluster appear to be largely
auto-dependent in their tripmaking, which each each exhibiting between
85-100\% of their modal split for car trips. This results in the parcels
within these zones having a large number of potentially accessible
locations in the travel time matrix. But by extension, the change in
accessibility over the pre- and during-COVID-19 time periods is affected
not only by the closure of service locations proximate to the zones, but
also the locations in the central city. The zone with the greatest
decrease in accessibility within this cluster (-0.0009) has a high rate
of car trips and connects to the most food bank locations in total as
well as those that stayed open or closed. The exception to this is the
zone in the middle of this cluster, which has a much higher proportion
of transit use 32\%. This modal split weighting results in fewer food
bank and service locations accessible from the parcels within this zone
which, by extension, limits their access to closed locations reachable
within 15 minutes in our calculations.

In the cluster to the south-west, the decrease in accessibility is
predominately driven by the closure of a high level-of-service Community
Meals provider. However, like the more central suburban zones,
low-income households within this cluster also between 90\% to 100\%
auto-dependent in their tripmaking in the TTS. The story is similar for
the zone located in the north-west that exhibits the greatest decrease
in accessibility. Here, low-income households responding to the TTS
conducted 100\% of their trips by car and, as a result, parcels within
this zone have access to the second-highest number of food bank and
service locations within 15 minutes. However, this also means zonal
accessibilities are greatly affected by the number of closures
throughout the city. Finally, in the city's north end and north-east
zones, low-income households exhibit a mixture of tripmaking behaviour
in the TTS. Households in some zones take transit more often and one
zone in particular has 100\% of its trips by walking while in others,
car use makes up 100\% of trips. For these zones, the decrease in
accessibility tends to be a product of the closure of several inner-city
food bank and service locations.

Just as the effects of the closures appear to have been uneven in space,
they also seem to have had different impacts on various population
segments. Table \ref{tab:accessibility-by-age} shows the estimated
number of people in each age group by their level of accessibility
before and during the pandemic. Here, it is important to note that the
quartiles are relative: people in the top 25\% of accessibility still
have lower accessibility during the pandemic than before. In reality,
every population group is worse off during the pandemic in terms of
their accessibility to food banks and services in the City of Hamilton.
However, some age groups were affected more. As seen in the table, the
distribution within quartiles for adults changed only slightly, and in
fact there are fewer adults in the bottom three quartiles as some found
themselves in the (deteriorated) top quartile of accessibility during
the pandemic. The story is different for seniors, a greater number of
whom are now in the third quartile, due to a combination of seniors
facing a worse accessibility situation, and just a few moving from the
bottom quartile (pre-pandemic) to the third during the pandemic. Those
aged 18 and less appear to have fared worse, and the largest change is
in the number of children who were in the third quartile before the
pandemic and found themselves in the bottom accessibility class during
the pandemic. These results suggest that, through a combination of the
typical modes of transportation of lower income households and the
spatial distribution of the population, the food bank/service locations
closed had a differential impact that more greatly affected the youngest
among the population in low income households.

@@

In the absence of information regarding how food insecure households
travel to food banks and related services, we examined accessibility to
food banks using a 15 minutes (one way) travel time threshold. The fact
that we must rely on a standard created for accessible drinking water in
the developing world only serves to highlight the tragedy of food
insecurity in an affluent country like Canada. More broadly, it points
to the absurd need to understand how a bad situation was made worse by
the pandemic: in effect, the analysis reveals that disparities in
accessibility to food banks and services predated the pandemic, that the
pandemic contributed to the deterioration of these services, and that
populations already in distress, particularly children, ended up in an
even more adverse state. How much worse, it is impossible to say, mainly
because there is also a dearth of information, let alone standards,
regarding what is an \emph{acceptable} or \emph{sufficient} level of
service when it comes to food banks. Whether this is one facility per
1,000 population or three facilities per 1,000 population, or any other
number, we know not.

@@

In terms of future research, it is important to note that the degree to
which our low-income cut-off of CAD\$40,000 reflects food insecurity in
the different zones of the study area is not known. We also consider all
food banks and food service locations equally. Information on
differential capacities at food provider locations is currently not
collected, and given the closure of facilities was not possible to
obtain. Data on services, such as the number of meals served, could be
used to refine the analysis in the future. Moreover, while the travel
survey allows us to model multi-modal accessibilities that align with
travel behaviour observed in the travel survey and capture differences
in accessibility by age categories, the use of travel survey data for
modelling food insecurity also has its limitations. Research into the
population weighting methods used in the TTS note that the survey may
under-count the lowest- and highest-income households in the survey
study region, although the magnitude of this under-counting is unknown,
and approximately 20\% of respondents to the 2016 survey did not report
their income (Rose, 2018). In that regard, the modal splits of
low-income households observed through the TTS data may not accurately
reflect the travel behaviours of food insecure populations, and our
estimates might be somewhat conservative if those households who do not
report income rely more on walking and/or transit for their mobility
needs.

>>>>

@@

Food insecurity is defined as an ``inadequate or uncertain access to a
sufficient quantity and/or adequate quality of food'' due to a
household's financial limitations (Enns et al., 2020). This condition
has been associated with reductions in nutritional outcomes
(Bhattacharya et al., 2004; Kirkpatrick and Tarasuk, 2008; Olson, 1999)
and negative physical and mental health impacts in children and adults
(Elgar et al., 2021; Jones, 2017; Ramsey et al., 2011; Seligman et al.,
2010; Stuff et al., 2004). Over at least the past four decades food
banks and related services have become an essential line of defense
against food insecurity in Canadian communities (Black and Seto, 2020;
Holmes et al., 2018; Riches, 2002; Tarasuk et al., 2020). In this
respect, Canada is not unlike numerous other wealthy countries where a
systematic dismantling of the welfare state took place in the
intervening period (Tarasuk et al., 2014).

@@

It is known that urban food environments, within which people make their
daily food choices, are essential in influencing eating behaviours and
health outcomes, based on factors such as food availability, ease of
geographic accessibility and socio-demographic variations (Paez et al.,
2010; Vanderlee and L'Abbé, 2017; Widener, 2018). However, while there
is a wealth of literature that has examined the topic of geographic
accessibility to healthy food through the ``food desert'' concept, there
has been little research into accessibility to food banks. Previous work
has explored differences in \emph{accessing} food banks, such as how
some households utilize food banks over short periods of time while
others regularly utilize food banks as longer-term resource (e.g., Enns
et al., 2020). In addition, transportation and locational considerations
have been raised as key issues in food bank accessibility in previous
qualitative research (Smith-Carrier et al., 2017). However, we are not
aware of any research that has focused on estimating or capturing this
geographic component of accessibility.

@@

For this study, we aim to look at how the landscape of food banks and
related services (e.g.~low-cost or free meal service providers)
available in Hamilton, Ontario, changed during the pandemic. Did the
number of open food bank services diminish? If so, what was the
accessibility to food banks before the pandemic from the perspective of
low income households, and how it changed during the pandemic? And
finally, who are most likely to have been impacted by changes in the
accessibility landscape? This paper first looks at the distribution of
food banks and related services before and during the pandemic. Then, we
use the balanced floating catchment area approach of Paez et al. (2019)
to investigate the accessibility situation. For this, we adopt a fully
disaggregated approach based on parcel-level data. Socio-economic and
demographic data are drawn from the latest Census of Canada (2016),
whereas travel information is from the most recent regional travel
survey from 2016. This paper follows reproducible research
recommendations (see Brunsdon and Comber, 2020), and the research was
conducted using open source tools for transportation analysis (Lovelace,
2021). The code and data necessary to reproduce the analysis are
available in a public repository\footnote{https://drive.google.com/drive/folders/1-l8hO1pMIqaBqf57j\_M\_BMrXhXFvrV2c?usp=sharing}.

@@

Food insecurity is the inability to acquire and consume an adequate
amount or good quality food, leading to inadequate nutrient intake (Enns
et al., 2020; Kirkpatrick and Tarasuk, 2008; Tarasuk and Vogt, 2009).
This nutrient deficiency is associated with major population health
concerns, particularly among Canadians at socio-economic disadvantage
(Bazerghi et al., 2016). Official government surveys such as the
Household Food Security Survey Module (HFSSM), the Canadian Community
Health Surveys (CCHS), the Longitudinal and International Study of
Adults (LISA), and official classifications determined by Health Canada
in relation to socio-demographic variables offer some insight into food
insecurity in Canada (Gundersen et al., 2018; Kirkpatrick and Tarasuk,
2008; Tarasuk and Vogt, 2009). Nationally, it has been found that food
insecurity impacts approximately 12.3\% of Canadian households (Tarasuk
et al., 2014)

@@

Economic stress caused by the COVID-19 pandemic, rising unemployment
rates, and changes in poverty levels have disrupted the food environment
and led to higher rates of food insecurity (Niles et al., 2020).
Pandemic-related food security studies in the US have found a
substantial increase in households experiencing food insecurity for the
first time, and also in households experiencing more severe food
insecurity than before (Niles et al., 2020; Wolfson and Leung, 2020).
Most recently in May 2020, Canada recorded 14.7\% of its population
living in food insecurity in the past 30 days (Statistics Canada,
2020a). Considering the negative mental and physical health effects,
increases in food insecurity rates due to the onset of COVID-19, signal
a change in the food environment with potential damages to population
health during the course of the pandemic and beyond (Niles et al.,
2020). Recent data\footnote{https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report\_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis\_Exec-Sum\_EN.pdf}
from FBC showed that 52\% of member food banks reported an increase in
usage in March of 2020 when initial lockdown restrictions were put in
place across much of the country. The pandemic also created significant
staffing issues with 42\% of food banks reporting a reduction in
volunteers. However, 53\% of food banks later reported a decrease in use
into the summer of 2020 which FBC members attributed to emergency
financial support programs from the federal government. Nevertheless,
some of these benefit programs were temporary. Although more recent
statistics on food bank use in Hamilton in 2020 and 2021 are not yet
available, data from the Daily Bread Food Bank in neighbouring Toronto
for August 2020 shows visits climbing 51\% year-over-year in that
city\footnote{https://www.dailybread.ca/wp-content/uploads/2020/11/Whos-Hungry-Report-2020.pdf},
which suggests that many households in Hamilton are likely to turn to
food bank services to meet their needs.

@@

Geographically, the ``old'' City of Hamilton was amalgamated with
several of its surrounding municipalities in 2001, with the city now
featuring a mix of urban, suburban, exurban, and rural areas. Lower-cost
housing proximate to the city's industrial north end has traditionally
attracted immigrants and less-affluent residents compared to the city's
wealthier suburbs. However, the decentralization of population from the
inner core has led to challenges in transit connectivity to amenities
and services and the proportion of auto users compared to transit users
remains very high (Behan et al., 2008; Topalovic et al., 2012). In
addition, the city is separated geographically by the Niagara
Escarpment. With sections of rocky cliff that approach 100m in height,
the escarpment presents a significant challenge for promoting active
travel and transport connections between ``mountain'' and ``lower city''
neighbourhoods. Taken together, the high level of food need, population
locations, and transportation network characteristics combine to inform
spatial accessibility to food banks and food services in the city.

@@

Figure \ref{fig:sensitivity-analysis} shows the results of conducting a
sensitivity analysis of the system-wide accessibility as we vary the
threshold (considering the situation before the pandemic). There is a
clear pattern whereby more strict values of \(\delta\) are associated
with higher levels of system-wide accessibility: while increases in
accessibility that result from decreases in the travel time window might
seem counter-intuitive, this is a result of lower \emph{congestion},
since fewer households are serviced and thus competition for the same
resources is more limited. System-wide accessibility declines with
higher values of \(\delta\): as more households are serviced, congestion
grows and the level of service declines, although this happens at a
declining rate. We are not aware of any research that explains how long
people are expected to travel for food banks, but we note that in
developing countries, accessible sources of drinking water are those
that can be reached in less than 30 minutes (round trip, see UNICEF-WHO,
2019). There is no reason why people in affluent countries should be
expected to spend more time travelling for a basic necessity such as
food. Accordingly, we adopt a 15-minute threshold for the analysis
(representing a one-way trip). This threshold is also approximately
where the rate of change in accessibility slows down.

@@

We find that the accessibility landscape deteriorated substantially
during the pandemic, with accessibility dropping on average by almost
41\%, but with large variations: some zones experienced changes in
accessibility of only about 10\%, whereas the most affected zone saw a
loss of accessibility of almost 71\%. Figure
\ref{fig:accessibility-changes-with-local-i} shows the changes in
accessibility. Every zone is worse off after the closure of facilities
during the pandemic, but some parts of the city seem to have been
particularly affected. To better highlight these changes, we used a
local indicator of spatial autocorrelation (Anselin, 1995) to explore
the pattern of change in accessibility. Twenty-four TAZs are flagged as
having significantly large losses of accessibility (at \(p\le 0.10\),
without correcting for multiple comparisons). Those zones are
highlighted in the figure, where it can be seen that they form more or
less compact neighborhoods. Remarkably, the largest significant drops in
accessibility are not downtown, but located in two cases in the
industrial north of the city, in one case in an inner suburb above the
escarpment, and lastly in a more suburban/exurban region in the
south-west.

@@

For the more suburban clusters of zones, the decrease in accessibility
is derived from the closure of locations throughout the city reachable
by car. In the cluster of central suburban zones for example, low-income
households in the outer ring of zones that exhibit medium to high
decreases in accessibility within this cluster appear to be largely
auto-dependent in their tripmaking, which each each exhibiting between
85-100\% of their modal split for car trips. This results in the parcels
within these zones having a large number of potentially accessible
locations in the travel time matrix. But by extension, the change in
accessibility over the pre- and during-COVID-19 time periods is affected
not only by the closure of service locations proximate to the zones, but
also the locations in the central city. The zone with the greatest
decrease in accessibility within this cluster (-0.0009) has a high rate
of car trips and connects to the most food bank locations in total as
well as those that stayed open or closed. The exception to this is the
zone in the middle of this cluster, which has a much higher proportion
of transit use (32\%). This modal split weighting results in fewer food
bank and service locations accessible from the parcels within this zone
which, by extension, limits their access to closed locations reachable
within 15 minutes in our calculations.

In the cluster to the south-west, the decrease in accessibility is
predominately driven by the closure of a high level-of-service Community
Meals provider. However, like the more central suburban zones,
low-income households within this cluster are also between 90\% to 100\%
auto-dependent in their tripmaking in the TTS. The story is similar for
the zone located in the north-west that exhibits the greatest decrease
in accessibility. Here, low-income households responding to the TTS
conducted 100\% of their trips by car and, as a result, dwellings within
this zone have access to the second-highest number of food bank and
service locations within 15 minutes. However, this also means zonal
accessibilities are greatly affected by the number of closures
throughout the city. Finally, in the city's north end and north-east
zones, low-income households exhibit a mixture of tripmaking behaviour
in the TTS. Households in some zones take transit more often and one
zone in particular has 100\% of its trips by walking while in others,
car use makes up 100\% of trips. For these zones, the decrease in
accessibility tends to be a product of the closure of several inner-city
food bank and service locations.

Just as the effects of the closures appear to have been uneven in space,
they also seem to have had different impacts on various population
segments. Using data on low income individuals by age drawn from the
TTS, Table \ref{tab:accessibility-by-age} shows the estimated number of
people in each age group by their level of accessibility before and
during the pandemic. Here, it is important to note that the quartiles
are relative: people in the top 25\% of accessibility still have lower
accessibility during the pandemic than before. In reality, every
population group is worse off during the pandemic in terms of their
accessibility to food banks and services in the City of Hamilton.
However, some age groups were affected more. As seen in the table, the
distribution within quartiles for adults changed only slightly, and in
fact there are fewer adults in the bottom three quartiles as some found
themselves in the (deteriorated) top quartile of accessibility during
the pandemic. The story is different for seniors, a greater number of
whom are now in the third quartile, due to a combination of seniors
facing a worse accessibility situation, and just a few moving from the
bottom quartile (pre-pandemic) to the third during the pandemic. Those
aged 18 and less appear to have fared worse, and the largest change is
in the number of children who were in the third quartile before the
pandemic and found themselves in the bottom accessibility class during
the pandemic. These results suggest that, through a combination of the
typical modes of transportation of lower income households and the
spatial distribution of the population, the food bank/service locations
closed had a differential impact that more greatly affected the youngest
among the population in low income households.

@@

In the absence of information regarding how food insecure households
travel to food banks and related services, we examined accessibility to
food banks using a 15 minutes (one-way) travel time threshold. The fact
that we must rely on a standard created for accessible drinking water in
the developing world only serves to highlight the tragedy of food
insecurity in an affluent country like Canada. More broadly, it points
to the absurd need to understand how a bad situation was made worse by
the pandemic: in effect, the analysis reveals that disparities in
accessibility to food banks and services predated the pandemic, that the
pandemic contributed to the deterioration of these services, and that
populations already in distress, particularly children, ended up in an
even more adverse state. How much worse, it is impossible to say, mainly
because there is also a dearth of information, let alone standards,
regarding what is an \emph{acceptable} or \emph{sufficient} level of
service when it comes to food banks. Whether this is one facility per
1,000 population or three facilities per 1,000 population, or any other
number, we know not. But like international human rights standards for
accessibility to clean drinking water within 30 minutes (round-trip),
this work strongly suggests that adopting better defined standards for
access to food banks that move beyond generalized conceptualizations of
``access'' to consider the geographic component of accessibility (and
its interactions with travel behaviour and individual/household
socioeconomic and demographic traits) would make a meaningful
contribution to understanding food access and ``food bank deserts'' for
food insecure populations.

@@

In terms of future research, it is important to note that the degree to
which our low-income cut-off of CAD\$40,000 reflects food insecurity in
the different zones of the study area is not known. We also consider all
food banks and food service locations equally. Information on
differential capacities at food provider locations is currently not
collected, and given the closure of facilities was not possible to
obtain. Data on services, such as the number of meals served, could be
used to refine the analysis in the future. Moreover, while the travel
survey allows us to model multi-modal accessibilities that align with
travel behaviour observed in the travel survey and capture differences
in accessibility by age categories, the use of travel survey data for
modelling food insecurity also has its limitations. Research into the
population weighting methods used in the TTS note that the survey may
under-count the lowest- and highest-income households in the survey
study region, although the magnitude of this under-counting is unknown,
and approximately 20\% of respondents to the 2016 survey did not report
their income (Rose, 2018). In that regard, the modal splits of
low-income households observed through the TTS data may not accurately
reflect the travel behaviours of food insecure populations, and our
estimates might in fact be somewhat conservative if those households who
do not report income rely more on walking and/or transit for their
mobility needs.
